Stochastic Approximation Monte Carlo with a Dynamic Update Factor (SAD)

### 2019-04-19 papers/histogram/sad.tex
cut hard-to-understand sentence
fixed a few references to figures in the paper.

Wang and Landau introduced an algorithm (WL) that uses an update
factor and a statistical histogram to compute the density of states of
a given system~\cite{wang2001determining, wang2001efficient}.  While
the method is incredibly powerful, it has a few disadvantage.
Firstly, it requires the user to select a number of parameters,
including the range of energies to be studied~\cite{landau2004new}.
This adds an additional hurdle to its application to a variety of
systems.  The simulation violates detailed balance, with the size of
the violation decreasing with time, which complicates convergence
analysis.  In fact, the error in a WL computation the error has been
demonstrated to saturate at a non-zero value~\cite{yan2003fast},
i.e. the method does not converge to the true density of
states~\cite{belardinelli2007wang, belardinelli2007fast,
  belardinelli2008analysis, belardinelli2014intrinsic,
  singh2012density}.  
%
Belardinelli and Pereyra demonstrated that allowing the update factor
to decrease faster than $1/t$ leads to
nonconvergence~\cite{belardinelli2007wang}.
%% Around the same time, the convergence rate for WL was being
%% explored~\cite{zhou2005understanding,lee2006convergence,
%%   belardinelli2007wang}.
%
This leads to their $1/t$-WL algorithm which ensures that the error
continues to decrease asymptotically as
$1/\sqrt{t}$~\cite{belardinelli2008analysis}. \davidsays{I don't
  understand the following: An effort to reduce the number of user
  defined parameters of WL and formulate a theory for why the method
  converges despite detailed balance being violated although
  infrequently was also being undertaken.  CUT THIS MAYBE?}

@@

A close examination of Fig.~\ref{fig:gamma-vs-t} shows that $\gamma$
for the Wang-Landau algorithm decreases much more rapidly than
$\nicefrac{1}{\sqrt{t}}$. It can be observed  for the simulations that
were carried out that all the methods from the weight-based family
struggle to converge if the update factor decreases too rapidlly.
Fig.~\ref{fig:N50-ff0.3-avg-error-slow} shows SAMC with inappropriate choices
of $t_0$ i.e. less than $\nicefrac{10^5}{\sqrt{t}}$ failing to converge
in a reasonable amount of time. Thus, the update factor $\gamma$, shown
as a function of simulation moves in Fig.~\ref{fig:gamma-vs-t}, proves
useful in determining which methods will converge in a reasonable
amount of time. We compare each method to a reference SAMC
simulation for an ideal choice of $t_0$.  This gives an unbiased
comparison with our developed method SAD. The dynamic version of SAMC
which we call SAD performs as well as SAMC (assuming the user can
correctly identify an appropriate choice of $t_0$ for SAMC).  SAD
outperforms WL and 1/t-WL? for the same given number of moves. The
absence of user defined parameters makes SAD an exciting alternative to
methods where user experience or successive implementation iteration
are needed to achieve reasonable results.

>>>>

Wang and Landau introduced an algorithm (WL) that uses an update
factor and a statistical histogram to compute the density of states of
a given system~\cite{wang2001determining, wang2001efficient}.  While
the method is incredibly powerful, it has a few disadvantage.
Firstly, it requires the user to select a number of parameters,
including the range of energies to be studied~\cite{landau2004new}.
This adds an additional hurdle to its application to a variety of
systems.  The simulation violates detailed balance, with the size of
the violation decreasing with time, which complicates convergence
analysis.  In fact, the error in a WL computation the error has been
demonstrated to saturate at a non-zero value~\cite{yan2003fast},
i.e. the method does not converge to the true density of
states~\cite{belardinelli2007wang, belardinelli2007fast,
  belardinelli2008analysis, belardinelli2014intrinsic,
  singh2012density}.  
%
Belardinelli and Pereyra demonstrated that allowing the update factor
to decrease faster than $1/t$ leads to
nonconvergence~\cite{belardinelli2007wang}.
%% Around the same time, the convergence rate for WL was being
%% explored~\cite{zhou2005understanding,lee2006convergence,
%%   belardinelli2007wang}.
%
This leads to their $1/t$-WL algorithm which ensures that the error
continues to decrease asymptotically as
$1/\sqrt{t}$~\cite{belardinelli2008analysis}.

@@

A close examination of Fig.~\ref{fig:gamma-vs-t} shows that $\gamma$
for the Wang-Landau algorithm decreases much more rapidly than
$\nicefrac{1}{\sqrt{t}}$. It can be observed  for the simulations that
were carried out that all the methods from the weight-based family
struggle to converge if the update factor decreases too rapidlly.
Fig.~\ref{fig:n256} shows SAMC with inappropriate choices
of $t_0$ i.e. less than $\nicefrac{10^4}{\sqrt{t}}$ failing to converge
in a reasonable amount of time. Thus, the update factor $\gamma$, shown
as a function of simulation moves in Fig.~\ref{fig:gamma-vs-t}, proves
useful in determining which methods will converge in a reasonable
amount of time. We compare each method to a reference SAMC
simulation for an ideal choice of $t_0$.  This gives an unbiased
comparison with our developed method SAD. The dynamic version of SAMC
which we call SAD performs as well as SAMC (assuming the user can
correctly identify an appropriate choice of $t_0$ for SAMC).  SAD
outperforms WL and 1/t-WL? for the same given number of moves. The
absence of user defined parameters makes SAD an exciting alternative to
methods where user experience or successive implementation iteration
are needed to achieve reasonable results.
